---
title: "US Traffic Accidents Analysis"
author: "Neelam Arshad"
execute: 
  echo: false
---


```{r}
#| message: false
#| results: hide
library(readr)     
library(ggplot2)    
library(lubridate)  

accidents <- read_csv("data/US_Accidents.csv", n_max = 500000)
```


```{r}
# Extract hour from Start_Time
accidents$Hour <- hour(accidents$Start_Time)

# Plot accidents by hour
ggplot(accidents, aes(x = Hour)) +
  geom_bar(fill = "#2C77B9", color = "white") +
  labs(title = "Number of Accidents by Hour of Day",
       x = "Hour of Day",
       y = "Number of Accidents") +
  theme_minimal(base_size = 14)
```

## Summary
Traffic accidents in the United States show distinct patterns based on the time of day. Using data from the US Accidents dataset, which includes detailed records of traffic incidents from 2016 to 2023, I analyzed 500,000 accident reports to explore temporal trends in accident frequency. The number of accidents rises sharply starting at 5 AM, peaks during the morning commute between 7 and 9 AM, and shows another increase in the late afternoon, reflecting typical workday travel patterns. This analysis reveals that daily human activity and traffic density play a significant role in when accidents are most likely to occur.

